---
date: "Annika created this code doc on 10/21/2023. Last compiled on `r Sys.Date()`"
output:
  github_document:
    df_print: kable
    toc: TRUE
    toc_depth: 1
---

# Description and Overview
This code summarizes the dataset that is a result of script: 10_mode-choice-cleaning

GitHub: This .Rmd code (possibly qmd if I have to change it with the new update to R Studio) is synced to Github, and when this code is knit, it's also synced to GitHub, as a .md file, which makes the knitted code easy to read (kind of like python code.) Currently in a FORK: <https://github.com/annitodd/GEMS-data/blob/main/gems-mode-choice>

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 
```

```{r}
#rm(list=ls()) # Clear RStudio environment
#cat("\014") # Clear console
```

## libraries

```{r}
library(arrow)
library(tidyverse)
library(readxl)
library(rstudioapi)
library(scales)
library(writexl)
```
## file path directories

```{r}
# get current root directory of the user's Github repo
root <- getwd() # Saves current WD 
#while ((basename(root) != "GEMS-data")) {
#  root <- dirname(root)
#} # Sets root equal to the location of the Github repo
#source(file.path(root, "paths.R")) # Runs paths.R file found in users Github repo
```

```{r}
data_path <- 'C:/FHWA/For FHWA folks/Mode_choice_estimation/Data'
data_results <- 'C:/FHWA_R2/mode_choice_estimation/data'
```
# Read in data
full merged dataset
```{r}
df_temp <- read_parquet(file.path(data_results, "10-mode-choice-cleaning_output-full-merged.parquet"))
names(df_temp)
```
recall that they don't equally merge, as we saw before (hhct has more households on file)
```{r}
df_temp |>
  count(rawdatafrom_trippub,rawdatafrom_tripct,rawdatafrom_hhct)
```

```{r}
df_temp <- ungroup(df_temp)
df_trips_only <- df_temp |> 
  filter(rawdatafrom_trippub==1,rawdatafrom_tripct==1,rawdatafrom_hhct==1)

df_trips_only |> 
  count(rawdatafrom_trippub,rawdatafrom_tripct,rawdatafrom_hhct)
df_trips_only <- ungroup(df_trips_only)
```
# Summary stats
## basic
how many trips in each state
```{r}
summary_table <- df_trips_only |> 
  group_by(ORIG_ST) |>
  summarise(   
    Ntrips = n(),
    .groups = "drop")
summary_table
```
how many trips for each mode
```{r}
summary_table <- df_trips_only |> 
  group_by(mode) |>
  summarise(   
    Ntrips = n(),
    .groups = "drop") |>
  arrange(-Ntrips)   
summary_table
```
## by county
how many trips in each county
choose only some vars
```{r}
df_trips_only_small <- df_trips_only |> 
  select(mode, trip_purpose, ORIG_COUNTRY,ORIG_ST,ORIG_CNTY)
df_trips_only_small %>% 
  distinct(mode,trip_purpose)
```
make wider dataset
```{r}
df_trips_only_small_wide <- df_trips_only_small  |>
  pivot_wider(
    names_from = mode,
    values_from = mode,
    values_fn = n_distinct
  )
df_trips_only_small_wide
```



```{r}
df_county_origin <- df_trips_only_small |>
  group_by(ORIG_ST,ORIG_CNTY) |>
  summarise(Ntrips = n(),
            across(contains("mode"),~n_distinct(.x),.names = "{.col}_Ndis"),
    .groups = "drop")
df_county_origin
```
```{r}
ggplot(df_county_origin,aes(x=Ntrips)) +
  geom_histogram(binwidth=0.5) +
  coord_cartesian(ylim = c(0, 50))
  

```

          across(where(is.numeric),~mean(.x),.names = "{.col}_avg"),
            across(where(is.factor ),~n_distinct(.x),.names = "{.col}_Ndis"),
 